---
title: "Pressing the (Dis)Advantage"
output:
  html_document:
    df_print: paged
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
---

```{r message=F, warning=F, echo=F}
library(reshape2)
library(dplyr)
library(ggplot2)
```

From [538 Riddler: Can You Find The Best Dungeons & Dragons Strategy?](https://fivethirtyeight.com/features/can-you-find-the-best-dungeons-dragons-strategy/)

> The fifth edition of Dungeons & Dragons introduced a system
> of “advantage and disadvantage.” When you roll a die “with
> advantage,” you roll the die twice and keep the higher result. Rolling
> “with disadvantage” is similar, except you keep the lower result
> instead. The rules further specify that when a player rolls with both
> advantage and disadvantage, they cancel out, and the player rolls a
> single die. Yawn!
> 
> There are two other, more mathematically interesting ways that advantage
> and disadvantage could be combined. First, you could have “advantage of
> disadvantage,” meaning you roll twice with disadvantage and then keep
> the higher result. Or, you could have “disadvantage of advantage,”
> meaning you roll twice with advantage and then keep the lower result. With
> a fair 20-sided die, which situation produces the highest expected
> roll: advantage of disadvantage, disadvantage of advantage or rolling
> a single die?
> 
> Extra Credit: Instead of maximizing your expected roll, suppose you
> need to roll N or better with your 20-sided die. For each value of N,
> is it better to use advantage of disadvantage, disadvantage of advantage
> or rolling a single die?

First I'll have a look at the conventional rolls: rolling a single die, rolling with advantage, rolling with disadvantage.

```{r}
iters <- 1000000
roll1 <- sample(1:20, iters, replace=T)
roll2 <- sample(1:20, iters, replace=T)
raw_rolls <- data.frame(advantage=pmax(roll1, roll2), single=roll1, disadvantage=pmin(roll1, roll2))
summary(raw_rolls)
```

The mean of the single roll is close to the expected 10.5. I'm not sure what I expect for advantage and disadvantage rolls, beyond the simple intuition that they are better/worse than a single roll, and that checks out here, with about 13.8 for advantage, 7.2 for disadvantage.

How about a visualization? Histograms of the simulation results approximate the probability mass functions.

```{r}
rolls <- melt(raw_rolls, id.vars=NULL)
names(rolls) <- c("type", "roll")
# re-order the factor so things are ordered left-to-right, best-to-worst
rolls$type <- factor(rolls$type, levels=c("advantage", "single", "disadvantage"))
roll_pmf <- rolls %>% group_by(type, roll) %>% summarize(count=n()) %>% mutate(frequency=count/sum(count))

p <- ggplot(roll_pmf, aes(x=roll, y=frequency, fill=type)) +
  geom_bar(stat="identity", show.legend=F) +
  scale_x_continuous(breaks=c(1, 5, 10, 15, 20))
p + facet_grid(. ~ type)
```

Interesting: the most likely result on an advantage roll is 20, and on a disadvantage roll it's 1.

Another way to look at the probabilties is with a boxplot. This shows the median, 25th, and 75th percentile. The expected value is marked with
a diamond.

```{r}
ggplot(rolls, aes(type, roll, color=type)) +
  geom_boxplot(show.legend=F) +
  stat_summary(fun.y=mean, geom="point", shape=18, size=3, show.legend=F) +
  scale_y_continuous(breaks=c(1, 5, 10, 15, 20))
```

Enough tinkering! For an advantage-of-disadvantage roll, you roll with disadvantage twice, and take the max. That's 4 single rolls. Disadvantage-of-advantage is the same, with disadvantage and advantage, min and max swapped.

I'll do 4 single rolls, and then reduce them down in two different ways to do advantage-of-disadvantage, disadvantage-of-advantage. I'll keep the first single roll for comparison against the other two distributions.

```{r}
roll1 <- sample(1:20, iters, replace=T)
roll2 <- sample(1:20, iters, replace=T)
roll3 <- sample(1:20, iters, replace=T)
roll4 <- sample(1:20, iters, replace=T)
raw_rolls <- data.frame(
  disadvantage_of_advantage=pmin(pmax(roll1, roll2), pmax(roll3, roll4)),
  single=roll1,
  advantage_of_disadvantage=pmax(pmin(roll1, roll2), pmin(roll3, roll4))
)
summary(raw_rolls)
```

Now I can answer the first question:

> With a fair 20-sided die, which situation produces the highest expected
> roll: advantage of disadvantage, disadvantage of advantage or rolling
> a single die?

Answer: disadvantage-of-advantage is best, with an expected value around 11.2. A single roll is the next best choice, with an expected value of 10.5. Last place goes to advantage-of-disadvantage, with an expected value around 9.8.

What do the approximate probability mass functions look like?

```{r}
rolls <- melt(raw_rolls, id.vars=NULL)
names(rolls) <- c("type", "roll")
# re-order the factor so things are ordered left-to-right, best-to-worst
rolls$type <- factor(rolls$type, levels=c("disadvantage_of_advantage", "single", "advantage_of_disadvantage"))

roll_pmf <- rolls %>% group_by(type, roll) %>% summarize(count=n()) %>% mutate(frequency=count/sum(count))

p <- ggplot(roll_pmf, aes(x=roll, y=frequency, fill=type)) +
  geom_bar(stat="identity", show.legend=F) +
  scale_x_continuous(breaks=c(1, 5, 10, 15, 20))
p + facet_grid(. ~ type)
```

Both disadvantage-of-advantage and advantage-of-disadvantage cluster in the middle, centered around 10 or so. If you're betting on rolling a 20 (or a 1), go with a single roll.

As a cross-check, boxplots confirm the answer, and highlight how much closer all three choices are to one another, compared to conventional advantage and disadvantage rolls.

```{r}
ggplot(rolls, aes(type, roll, color=type)) +
  geom_boxplot(show.legend=F) +
  stat_summary(fun.y=mean, geom="point", shape=18, size=3, show.legend=F) +
  scale_y_continuous(breaks=c(1, 5, 10, 15, 20))
```

How about that extra credit question?

> Instead of maximizing your expected roll, suppose you
> need to roll N or better with your 20-sided die. For each value of N,
> is it better to use advantage of disadvantage, disadvantage of advantage
> or rolling a single die?

The simulated probability mass functions answer this one. For example, if we need to roll 19 or higher, the odds of success are the probability of rolling a 19, plus the odds of rolling a 20, for each of the three roll types. We'll calculate all the odds with a cumulative sum, from 20 downward.

Sanity checks/expectations: for each roll type, we expect to see 100% success if N=1. As N increases, the odds of success decrease, down to the minimum at 20 (the odds of getting 20, for each type).

```{r}
roll_pmf <- roll_pmf %>% mutate(rcdf = rev(cumsum(rev(frequency))))
p <- ggplot(roll_pmf, aes(x=roll, y=rcdf, color=type)) +
  geom_point() +
  geom_line() +
  scale_x_continuous(breaks=c(1, 5, 10, 15, 20))
p + xlab("N") + ylab("odds of rolling N or higher")
```

Advantage-of-disadvantage is only an optimal choice when N=1 (you need to roll 1 or higher - not an interesting situation!). It's vacuous: all 3 choices give the same "odds". Otherwise, choose disadvantage-of-advantage for N up to 13, and choose single roll for N=14 and higher.

Source code is [here](https://github.com/aaronferrucci/dnd_vantage)
Github hosting is [here](https://aaronferrucci.github.io/dnd_vantage/)

